---
title: "Forecasting Public Transport Demand"
author: "Miguel Conde"
date: "19 de abril de 2017"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

## Problem Description

* A transportation company is expecting increased demand for its services.
* In consequence, this company is planning certain investments to satisfy the
adtional demand.
* We need a reliable forecast of future demand to quantify the needed investment.
* To create such forecast we have at disposal historical data, on each 15-min
interval between 6:30 and 22:00, on the number of arriving passengers.

Our aim is to create a forecasting method that can generate orecasts for the number of passengers arriving.

## Data
The historic data contains demand for a 3-week period (2005-03-01 to 2005-03-21), separated in 15-min intervals.
```{r}
library(readxl)
file_n <- file.path(".", "DATA", "bicup2006.xls")
bicup2006 <- read_excel(file_n, skip = 5,
                        col_names = c("", "", "", "DATE", "TIME", "DEMAND"),
                        col_types = c("blank", "blank", "blank", 
                                      "date", "date", "numeric"))
library(lubridate)
bicup2006$DT <- as.POSIXct(paste(bicup2006$DATE, sprintf("%d:%d:%d", 
                                                  hour(bicup2006$TIME),
                                                  minute(bicup2006$TIME),
                                                  second(bicup2006$TIME))))
bicup2006 <- data.frame(DT = bicup2006$DT, DEMAND = bicup2006$DEMAND)
```

```{r}
library(lubridate)
n_days <- max(day(bicup2006$DT)) - min(day(bicup2006$DT)) + 1
n_hours <- interval(min(bicup2006$DT) + days(n_days - 1), # From 06:30 to 22:00
                    max(bicup2006$DT) + minutes(15)) / hours(1)
n_quarters <- 4*n_hours # From (06:30 - 06:45) to (22:00 - 22:15)

bicup2006_ts <- ts(bicup2006$DEMAND, start = c(1, 1), frequency = n_quarters)
plot(bicup2006_ts, ylab = "Number of passengers", xlab = "Time",
     main = "Transport Demand", xaxt = "n")
axis(1, at = seq(1, n_days, 1), 
     labels = format(as.Date(bicup2006$DT[n_quarters*seq(1, n_days, 1)])),
     las = 3, cex.axis = .5)
```

## Goal
Training period <- 1st 2 weeks
Validation period <- last week

```{r}
startDate_total <- min(bicup2006$DT)
endDate_total   <- max(bicup2006$DT)

startDate_train <- min(bicup2006$DT)
endDate_train   <- endDate_total - days(7)

startDate_valid <- startDate_train + days(14)
endDate_valid   <- endDate_total
```


```{r}
nTrain <- 14*n_quarters
nTotal <- length(bicup2006_ts)
nValid <- nTotal - nTrain

startTrain <- start(bicup2006_ts)
endTrain <- startTrain + c(0, nTrain - 1)
startValid <- endTrain + c(0, 1)
endValid <- startValid + c(0, nValid - 1)

bicup2006_train_ts <- window(bicup2006_ts, start = startTrain, end = endTrain)
bicup2006_valid_ts <- window(bicup2006_ts, start = startValid, end = endValid)
```

Our goal is to create a model that generates reasonably good forecasts on any
time/day of the week.

Consider not only metrics as MAE, MAPE and RMSE, but also look at the actual and forecasted values overlaid on a time plot.

## Assignment


## Tips and Suggested Steps

### 1 Exploratory analysis

To identify the components of this ts. Is there a trend? Is there seasonality?
If so, how many seasons are there? Are there any other visible patterns? Are the patterns global (the same throughout the series) or local?


```{r}
library(forecast)
Acf(bicup2006_ts, lag.max = 630)
```


```{r}
daily_bicup2006 <- ts(bicup2006$DEMAND, start = c(1, 1), 
                      frequency = n_quarters)
plot(daily_bicup2006,
     main = "Daily periodicity")
```

```{r}
plot(stl(daily_bicup2006, s.window = "periodic"),
     main = "Daily periodicity")
```


```{r}
weekly_bicup2006 <- ts(bicup2006$DEMAND, start = c(1, 1), 
                       frequency = n_quarters*7)
plot(weekly_bicup2006,
     main = "Weekly periodicity")
```

```{r}
plot(stl(weekly_bicup2006, s.window = "periodic"),
     main = "Weekly periodicity")
```


There are **two sesonalities**: **weekly (period 7 days)** and **daily (24 hours before)**. 


#### Predictability


## Solutions

### Naive forecasts

### Double seasoned holt-winter's model

```{r}
train_msts <- msts(bicup2006[bicup2006$DT <= endDate_train, "DEMAND"],
                   seasonal.periods = n_quarters * c(1, 7),
                   start = startTrain)
VALID_msts <- msts(bicup2006[bicup2006$DT > endDate_train, "DEMAND"],
                   seasonal.periods = n_quarters * c(1, 7),
                   start = startValid)

# Fit a double seasoned holt-winter's model
bicup2006_dshw_pred <- dshw(train_msts+1, h = nValid)
# This is to correct a minor error in forecast package v7.1
bicup2006_dshw_pred_mean <- msts(bicup2006_dshw_pred$mean-1,
                                   seasonal.periods = n_quarters * c(1, 7),
                                   start = startValid)
accuracy(bicup2006_dshw_pred_mean, VALID_msts)
```

```{r}
plot(train_msts, 
     xlim = c(1, 4),
     ylab = "Number of passengers", xlab = "Time",
     main = "Transport Demand", xaxt = "n")
axis(1, at = (seq(1, n_days+7, 7)-1)/7+1, 
     labels = format(as.Date(bicup2006$DT[n_quarters*seq(1, n_days+7, 7)])),
     las = 3, cex.axis = .75)
lines(bicup2006_dshw_pred_mean, lwd = 2, col = "blue")
```


#### With 2nd-order level forecasting model on the residual series
```{r}
train_res_arima <- Arima(residuals(bicup2006_dshw_pred),
                         order = c(1, 0, 0))
summary(train_res_arima)
train_res_arima_pred <- forecast(train_res_arima)
```
```{r}
plot(residuals(bicup2006_dshw_pred), 
     xlim = c(1, 4),
     ylab = "Residuals", xlab = "Time", xaxt = "n")
axis(1, at = (seq(1, n_days+7, 7)-1)/7+1, 
     labels = format(as.Date(bicup2006$DT[n_quarters*seq(1, n_days+7, 7)])),
     las = 3, cex.axis = .75)

Acf(residuals(train_res_arima), main = "", lwd = 5, 
    col = "lightblue")
```

```{r}
# Now we have the model, let's make sonme forecasting
# 1st level model forecasting is bicup2006_dshw_pred_mean
plot(train_msts, 
     xlim = c(1, 4), lwd = 2,
     ylab = "Number of passengers", xlab = "Time",
     main = "Transport Demand", xaxt = "n")
axis(1, at = (seq(1, n_days+7, 7)-1)/7+1, 
     labels = format(as.Date(bicup2006$DT[n_quarters*seq(1, n_days+7, 7)])),
     las = 3, cex.axis = .75)
lines(VALID_msts, lwd = 2, lty = 2, col = "black")

# 2nd order level model, 1st level model + AR(1) model (forecasting = train_res_arima_pred)

fitted_v <- fitted(bicup2006_dshw_pred) + fitted(train_res_arima_pred)
fitted_m <- bicup2006_dshw_pred_mean + train_res_arima_pred$mean

lines(fitted_v, lwd = 1, col = "blue")
lines(fitted_m, lwd = 1, col = "blue", lty = 2)
```

```{r}
# 1st order accuracy
accuracy(fitted(bicup2006_dshw_pred), train_msts)
accuracy(bicup2006_dshw_pred_mean, VALID_msts)
```

```{r}
# 2nd order accuracy
accuracy(fitted_v, train_msts)
accuracy(fitted_m, VALID_msts)
```

### Neural networks

Approach:

* Daily seasonality: lag-(n_quarters)
* Weekly seasonality: 7 dummy variables

```{r}
bicup2006$day_of_week <- wday(bicup2006$DT, label = TRUE)

day_of_week.dummies <- model.matrix(~ 0 + day_of_week, 
                                    data = bicup2006)
colnames(day_of_week.dummies) <- gsub("day_of_week", "",
                                      colnames(day_of_week.dummies))

bicup2006$lag_n_quarters <- c(rep(NA, n_quarters),
                              bicup2006$DEMAND[1:(nrow(bicup2006)-n_quarters)])
bicup2006 <- cbind(bicup2006, day_of_week.dummies[,-1])
```




#### nnetar()
```{r}
# External regressors names
xreg_features <- c(names(bicup2006)[5:10])
xreg_train <- bicup2006[bicup2006$DT <= endDate_train, xreg_features]
xreg_valid <- bicup2006[bicup2006$DT > endDate_train, xreg_features]

# Weekly seasonality
train_ts <- ts(bicup2006[bicup2006$DT <= endDate_train, "DEMAND"],
                   start = startTrain, frequency = 7*n_quarters)
valid_ts <- ts(bicup2006[bicup2006$DT > endDate_train, "DEMAND"],
               start = startValid, frequency = 7*n_quarters)

# Daily seasonality: n_quarters non seasonal lags 
# Weekly seasonality: 6 dummy variables and 1 seasonal (weekly) lag
set.seed(201)
bicup2006.nnetar <- nnetar(train_ts, repeats = 20, 
                           p = n_quarters, P = 1, size = 4,
                           xreg = xreg_train)

bicup2006.nnetar.pred <- forecast(bicup2006.nnetar, 
                                  h = nValid,
                                  xreg = xreg_valid)
accuracy(bicup2006.nnetar.pred, valid_ts)
```

```{r}
plot(bicup2006.nnetar.pred)
lines(train_ts, lty = 2, col = "red")
lines(valid_ts, lty = 2, col = "red")
```



#### elman


## Others

### TMA and Simple Exponential Smoothing
MA and SES can be used to forecast series with just a level and noise.
If the series shows, in addition, a trend and/or seasonality, we can first 
remove those components.

MA and SES are not valid with trend or seasonality. First, we should de-trend and deseasonalize using Regression models, Advanced exponential smoothing or differencing

### Advanced Exponential Smoothing
AES can capture trend and seasonality.
```{r}
# Weekly seasonality
train_ts <- ts(bicup2006[bicup2006$DT <= endDate_train, "DEMAND"],
                   start = startTrain, frequency = 7*n_quarters)
valid_ts <- ts(bicup2006[bicup2006$DT > endDate_train, "DEMAND"],
               start = startValid, frequency = 7*n_quarters)

bicup2006_ets <- ets(y = train_ts)
```

```{r}
bicup2006_ets_pred <- forecast(bicup2006_ets, h = nValid, level = 0)

plot(bicup2006_ets_pred,
     xlab = "Time", ylab = "Number of passengers", main = "Transport Demand",
     flty = 2)

lines(bicup2006_ets_pred$fitted, lwd = 2, col = "blue")
lines(valid_ts)

accuracy(bicup2006_ets_pred, valid_ts)

```

### REGRESSION-BASED MODELS - tslm()
```{r}
# Weekly seasonality
train_ts <- ts(bicup2006[bicup2006$DT <= endDate_train, "DEMAND"],
                   start = startTrain, frequency = 7*n_quarters)
valid_ts <- ts(bicup2006[bicup2006$DT > endDate_train, "DEMAND"],
               start = startValid, frequency = 7*n_quarters)

bicup2006_lm <- tslm(train_ts ~ trend + season)
```

```{r}
bicup2006_lm_pred <- forecast(bicup2006_lm, h = nValid, level = 0)

plot(bicup2006_lm_pred,
     xlab = "Time", ylab = "Number of passengers", main = "Transport Demand",
     flty = 2)

lines(bicup2006_lm_pred$fitted, lwd = 2, col = "blue")
lines(valid_ts)

accuracy(bicup2006_lm_pred, valid_ts)

```


### ARIMA models

```{r}
bicup2006_arima <- auto.arima(train_ts, xreg = xreg_train)

# bicup2006_arima <- Arima(train_ts, 
#                          order = c(4, 1, 2), 
#                          seasonal = list(order = c(2,1,1), 
#                                          period = n_quarters), 
#                          xreg = xreg_train)
```

```{r}
bicup2006_arima_pred <- forecast(bicup2006_arima, h = nValid, level = 0,
                                 xreg = xreg_valid)

plot(bicup2006_arima_pred,
     xlab = "Time", ylab = "Number of passengers", main = "Transport Demand",
     flty = 2)

lines(bicup2006_arima_pred$fitted, lwd = 2, col = "blue")
lines(valid_ts)

accuracy(bicup2006_arima_pred, valid_ts)

```

### 1st and 2nd order models

