---
title: "timekit package"
author: "Miguel Conde"
date: "30 de mayo de 2017"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

https://business-science.github.io/timekit/index.html


The package contains the following functions:

* **Get an index**: `tk_index` returns the time series index of time series objects, models. The argument `timekit_idx` can be used to return a special timekit “index” attribute for regularized ts objects that returns a non-regularized date / date-time index if present.

* **Get critical timeseries information**: `tk_get_timeseries_signature` and `tk_get_timeseries_summary` takes an index and provides a time series decomposition and key summary attributes of the index, respectively. The `tk_augment_timeseries_signature` expedites adding the time series decomposition to the time series object.

* **Make a future timeseries**: `tk_make_future_timeseries` models a future time series after an existing time series index.

* **Coercion functions**: `tk_tbl`, `tk_ts`, `tk_xts`, `tk_zoo`, and `tk_zooreg` coerce time-based tibbles tbl to and from each of the main time-series data types `xts`, `zoo`, `zooreg`, `ts`, maintaining the time-based index.


# Time Series Coercion

```{r}
require(timekit)
require(tidyquant)

ten_year_treasury_rate_tbl <- tq_get("DGS10", 
                                     get  = "economic.data", 
                                     from = "1997-01-01", 
                                     to   = "2016-12-31") %>%
  rename(pct = price) %>%
  mutate(pct = pct / 100)

ten_year_treasury_rate_tbl
```

```{r}
class(ten_year_treasury_rate_tbl)
```

```{r}
str(ten_year_treasury_rate_tbl)
```

```{r}
# Change to a quarterly periodicity
ten_year_treasury_rate_tbl <- ten_year_treasury_rate_tbl %>%
  tq_transmute(pct, mutate_fun = to.period, period = "quarters")

ten_year_treasury_rate_tbl
```

## Coercion issues with `ts()`
The `ts` data structure is the most difficult to coerce back and forth because by default it does not contain a time-based index. Rather it uses a regularized index computed using the `start` and `frequency` arguments. Coercion to `ts` is done using the `ts()` function from the `stats` library, which results in various problems.

```{r}
# date column gets coerced to numeric
ts(ten_year_treasury_rate_tbl, start = 1997, freq = 4) %>%
    head()
```
The correct method is to call the specific column desired. However, this presents a new issue. The date index is lost, and a different “regularized” index is built using the start and frequency attributes.

```{r}
ten_year_treasury_rate_ts_stats <- ts(ten_year_treasury_rate_tbl[,"pct"], 
                                      start = 1997, 
                                      freq  = 4)
ten_year_treasury_rate_ts_stats
```

```{r}
# No date index attribute
str(ten_year_treasury_rate_ts_stats)
```

We can get the index using the `index()` function from the `zoo` package. The index retained is a regular sequence of numeric values. In many cases, the regularized values cannot be coerced back to the original time-base because the date and date time data contains significantly more information (i.e. year-month-day, hour-minute-second, and timezone attributes) and the data may not be on a regularized interval (frequency).
```{r}
# Regularized numeric sequence
index(ten_year_treasury_rate_ts_stats)
```

**Solution:**

```{r}
# date automatically dropped and user is warned
ten_year_treasury_rate_ts_timekit <- tk_ts(ten_year_treasury_rate_tbl, 
                                         start = 1997, 
                                         freq  = 4)
```

```{r}
ten_year_treasury_rate_ts_timekit
```

```{r}
# More attributes including time index, time class, time zone
str(ten_year_treasury_rate_ts_timekit)
```

**`tk_index()`** and **`tk_tbl()`**
```{r}
# Can now retrieve the original date index
timekit_index <- tk_index(ten_year_treasury_rate_ts_timekit, timekit_idx = TRUE)
head(timekit_index)
```
```{r}
class(timekit_index)
```

Note the difference:
```{r}
tk_index(ten_year_treasury_rate_ts_timekit, timekit_idx = FALSE)
```
```{r}
# Coercion back to tibble using the default index (regularized)
ten_year_treasury_rate_ts_timekit %>%
    tk_tbl(index_rename = "date", timekit_idx = FALSE)
```
Compare to this (the original index now):
```{r}
# Coercion back to tibble now using the timekit index (date / date-time)
ten_year_treasury_rate_tbl_timekit <- ten_year_treasury_rate_ts_timekit %>%
    tk_tbl(index_rename = "date", timekit_idx = TRUE)
ten_year_treasury_rate_tbl_timekit
```

```{r}
# Comparing the coerced tibble with the original tibble
identical(ten_year_treasury_rate_tbl_timekit, ten_year_treasury_rate_tbl)
```

## Coercion methods

## Additional Concepts

### Testing if an object has a timekit index

### Coercing ts to xts and zoo

### Working with yearmon and yearqtr index

### Getting the index of other time-based objects (e.g. models)


# Working with the Time Series Index

The **time series index**, which consists of a collection of time-based values that define when each observation occurred, is the most important part of a time series object. 

The index gives the user a lot of information in a simple timestamp. Consider the datetime “2016-01-01 00:00:00”: 

* From this timestamp, we can decompose the datetime to get the **signature**, which consists of the year, quarter, month, day, day of year, day of month, hour, minute, and second of the occurrence of a single observation. 

* Further, the difference between two or more observations is the **frequency** from which we can obtain even more information such as the periodicity of the data and whether or not these observations are on a regular interval. This information is critical as it provides the basis for performance over time in finance, decay rates in biology, growth rates in economics, and so on.

```{r}
FB_tbl <- FANG %>% filter(symbol == "FB")
FB_tbl
```

```{r}
FB_vol_date <- FB_tbl %>% select(date, volume)
FB_vol_date
```

```{r}
FB_vol_yearqtr <- FB_vol_date %>%
    mutate(date = as.yearqtr(date)) %>%
    group_by(date) %>%
    summarize(volume = sum(volume))
FB_vol_yearqtr
```

## Extract an index

**`tk_index`**
The function tk_index() extracts the index from any time series object including data frame (or tbl), xts, zoo, etc. 

```{r}
# idx_date
idx_date <- tk_index(FB_vol_date)
str(idx_date)
```

```{r}
# idx_yearqtr
idx_yearqtr <- tk_index(FB_vol_yearqtr)
paste0("class: ", class(idx_yearqtr), "\n",
       "n.obs: ", length(idx_yearqtr), "\n",
       "head:  ", stringr::str_c(head(idx_yearqtr), collapse = ", ")) %>%
    cat()
```


## Analyze the index

### tk_get_timeseries_signature and tk_augment_timeseries_signature

The index can be decomposed into a *signature*. The time series signature is a unique set of properties of the time series values that describe the time series. The function `tk_get_timeseries_signature()` can be used to convert the index to a tibble containing the relevant values (columns):
```{r}
# idx_date signature
tk_get_timeseries_signature(idx_date)
```
We can also get the signature of other time-based objects. The `tk_get_timeseries_signature()` function works with date, datetime, yearmon, and yearqtr vectors.

```{r}
# idx_yearqtr signature
tk_get_timeseries_signature(idx_yearqtr)
```
It’s usually important to keep the index signature with the values (e.g. volume in our example). We can use an expedited approach with `tk_augment_timeseries_signature()`, which adds the signature to the end of the time series object.
```{r}
# Augmenting a data frame
FB_vol_date_signature <- tk_augment_timeseries_signature(FB_vol_date)
FB_vol_date_signature
```
Analyzing the time series is now easier with the decomposed measures. For example, we can create a month plot very easily in `ggplot2`.

```{r}
# Example Benefit 1: Making a month plot
FB_vol_monthly <- FB_vol_date_signature %>%
    group_by(year, month.lbl) %>%
    summarize(volume = sum(volume)) 

FB_vol_monthly %>%
    ggplot(aes(x = month.lbl, y = volume, fill = factor(year))) +
    geom_bar(stat = "identity") +
    labs(title = "Month Plot of FB Volume", x ="", fill = "Year",
         subtitle = "Analyzing time-based metrics is easy with time series signature") +
    theme_tq() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_tq() +
    scale_y_continuous(labels = scales::comma)
```

Modeling is also much easier. As an example, we can use linear regression model using the lm() function with the month and year as a predictor of volume.

```{r}
# Example Benefit 2: Modeling is easier
fit <- lm(volume ~ year + month.lbl, data = FB_vol_monthly)
summary(fit)
```
### tk_get_timeseries_summary
```{r}
# idx_date: First six columns, general summary
tk_get_timeseries_summary(idx_date)[,1:6]
```


```{r}
# idx_date: First six columns, differences summary
tk_get_timeseries_summary(idx_date)[,7:12]
```

```{r}
# idx_yearqtr: First six columns, general summary
tk_get_timeseries_summary(idx_yearqtr)[,1:6]
```

```{r}
# idx_yearqtr: Last six columns, difference summary
tk_get_timeseries_summary(idx_yearqtr)[,7:12]
```



# Making a Future Time Series Index

# Forecasting Using a Time Series Signature