---
title: "timekit package"
author: "Miguel Conde"
date: "30 de mayo de 2017"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
```

https://business-science.github.io/timekit/index.html


The package contains the following functions:

* **Get an index**: `tk_index` returns the time series index of time series objects, models. The argument `timekit_idx` can be used to return a special timekit “index” attribute for regularized ts objects that returns a non-regularized date / date-time index if present.

* **Get critical timeseries information**: `tk_get_timeseries_signature` and `tk_get_timeseries_summary` takes an index and provides a time series decomposition and key summary attributes of the index, respectively. The `tk_augment_timeseries_signature` expedites adding the time series decomposition to the time series object.

* **Make a future timeseries**: `tk_make_future_timeseries` models a future time series after an existing time series index.

* **Coercion functions**: `tk_tbl`, `tk_ts`, `tk_xts`, `tk_zoo`, and `tk_zooreg` coerce time-based tibbles tbl to and from each of the main time-series data types `xts`, `zoo`, `zooreg`, `ts`, maintaining the time-based index.


# Time Series Coercion

```{r}
require(timekit)
require(tidyquant)

ten_year_treasury_rate_tbl <- tq_get("DGS10", 
                                     get  = "economic.data", 
                                     from = "1997-01-01", 
                                     to   = "2016-12-31") %>%
  rename(pct = price) %>%
  mutate(pct = pct / 100)

ten_year_treasury_rate_tbl
```

```{r}
class(ten_year_treasury_rate_tbl)
```

```{r}
str(ten_year_treasury_rate_tbl)
```

```{r}
# Change to a quarterly periodicity
ten_year_treasury_rate_tbl <- ten_year_treasury_rate_tbl %>%
  tq_transmute(pct, mutate_fun = to.period, period = "quarters")

ten_year_treasury_rate_tbl
```

## Coercion issues with `ts()`
The `ts` data structure is the most difficult to coerce back and forth because by default it does not contain a time-based index. Rather it uses a regularized index computed using the `start` and `frequency` arguments. Coercion to `ts` is done using the `ts()` function from the `stats` library, which results in various problems.

```{r}
# date column gets coerced to numeric
ts(ten_year_treasury_rate_tbl, start = 1997, freq = 4) %>%
    head()
```
The correct method is to call the specific column desired. However, this presents a new issue. The date index is lost, and a different “regularized” index is built using the start and frequency attributes.

```{r}
ten_year_treasury_rate_ts_stats <- ts(ten_year_treasury_rate_tbl[,"pct"], 
                                      start = 1997, 
                                      freq  = 4)
ten_year_treasury_rate_ts_stats
```

```{r}
# No date index attribute
str(ten_year_treasury_rate_ts_stats)
```

We can get the index using the `index()` function from the `zoo` package. The index retained is a regular sequence of numeric values. In many cases, the regularized values cannot be coerced back to the original time-base because the date and date time data contains significantly more information (i.e. year-month-day, hour-minute-second, and timezone attributes) and the data may not be on a regularized interval (frequency).
```{r}
# Regularized numeric sequence
index(ten_year_treasury_rate_ts_stats)
```

**Solution:**

```{r}
# date automatically dropped and user is warned
ten_year_treasury_rate_ts_timekit <- tk_ts(ten_year_treasury_rate_tbl, 
                                         start = 1997, 
                                         freq  = 4)
```

```{r}
ten_year_treasury_rate_ts_timekit
```

```{r}
# More attributes including time index, time class, time zone
str(ten_year_treasury_rate_ts_timekit)
```

**`tk_index()`** and **`tk_tbl()`**
```{r}
# Can now retrieve the original date index
timekit_index <- tk_index(ten_year_treasury_rate_ts_timekit, timekit_idx = TRUE)
head(timekit_index)
```
```{r}
class(timekit_index)
```

Note the difference:
```{r}
tk_index(ten_year_treasury_rate_ts_timekit, timekit_idx = FALSE)
```
```{r}
# Coercion back to tibble using the default index (regularized)
ten_year_treasury_rate_ts_timekit %>%
    tk_tbl(index_rename = "date", timekit_idx = FALSE)
```
Compare to this (the original index now):
```{r}
# Coercion back to tibble now using the timekit index (date / date-time)
ten_year_treasury_rate_tbl_timekit <- ten_year_treasury_rate_ts_timekit %>%
    tk_tbl(index_rename = "date", timekit_idx = TRUE)
ten_year_treasury_rate_tbl_timekit
```

```{r}
# Comparing the coerced tibble with the original tibble
identical(ten_year_treasury_rate_tbl_timekit, ten_year_treasury_rate_tbl)
```

## Coercion methods

## Additional Concepts

### Testing if an object has a timekit index

### Coercing ts to xts and zoo

### Working with yearmon and yearqtr index

### Getting the index of other time-based objects (e.g. models)


# Working with the Time Series Index

The **time series index**, which consists of a collection of time-based values that define when each observation occurred, is the most important part of a time series object. 

The index gives the user a lot of information in a simple timestamp. Consider the datetime “2016-01-01 00:00:00”: 

* From this timestamp, we can decompose the datetime to get the **signature**, which consists of the year, quarter, month, day, day of year, day of month, hour, minute, and second of the occurrence of a single observation. 

* Further, the difference between two or more observations is the **frequency** from which we can obtain even more information such as the periodicity of the data and whether or not these observations are on a regular interval. This information is critical as it provides the basis for performance over time in finance, decay rates in biology, growth rates in economics, and so on.

```{r}
FB_tbl <- FANG %>% filter(symbol == "FB")
FB_tbl
```

```{r}
FB_vol_date <- FB_tbl %>% select(date, volume)
FB_vol_date
```

```{r}
FB_vol_yearqtr <- FB_vol_date %>%
    mutate(date = as.yearqtr(date)) %>%
    group_by(date) %>%
    summarize(volume = sum(volume))
FB_vol_yearqtr
```

## Extract an index

**`tk_index`**
The function tk_index() extracts the index from any time series object including data frame (or tbl), xts, zoo, etc. 

```{r}
# idx_date
idx_date <- tk_index(FB_vol_date)
str(idx_date)
```

```{r}
# idx_yearqtr
idx_yearqtr <- tk_index(FB_vol_yearqtr)
paste0("class: ", class(idx_yearqtr), "\n",
       "n.obs: ", length(idx_yearqtr), "\n",
       "head:  ", stringr::str_c(head(idx_yearqtr), collapse = ", ")) %>%
    cat()
```


## Analyze the index

### tk_get_timeseries_signature and tk_augment_timeseries_signature

The index can be decomposed into a *signature*. The time series signature is a unique set of properties of the time series values that describe the time series. The function `tk_get_timeseries_signature()` can be used to convert the index to a tibble containing the relevant values (columns):
```{r}
# idx_date signature
tk_get_timeseries_signature(idx_date)
```
We can also get the signature of other time-based objects. The `tk_get_timeseries_signature()` function works with date, datetime, yearmon, and yearqtr vectors.

```{r}
# idx_yearqtr signature
tk_get_timeseries_signature(idx_yearqtr)
```
It’s usually important to keep the index signature with the values (e.g. volume in our example). We can use an expedited approach with `tk_augment_timeseries_signature()`, which adds the signature to the end of the time series object.
```{r}
# Augmenting a data frame
FB_vol_date_signature <- tk_augment_timeseries_signature(FB_vol_date)
FB_vol_date_signature
```
Analyzing the time series is now easier with the decomposed measures. For example, we can create a month plot very easily in `ggplot2`.

```{r}
# Example Benefit 1: Making a month plot
FB_vol_monthly <- FB_vol_date_signature %>%
    group_by(year, month.lbl) %>%
    summarize(volume = sum(volume)) 

FB_vol_monthly %>%
    ggplot(aes(x = month.lbl, y = volume, fill = factor(year))) +
    geom_bar(stat = "identity") +
    labs(title = "Month Plot of FB Volume", x ="", fill = "Year",
         subtitle = "Analyzing time-based metrics is easy with time series signature") +
    theme_tq() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_tq() +
    scale_y_continuous(labels = scales::comma)
```

Modeling is also much easier. As an example, we can use linear regression model using the lm() function with the month and year as a predictor of volume.

```{r}
# Example Benefit 2: Modeling is easier
fit <- lm(volume ~ year + month.lbl, data = FB_vol_monthly)
summary(fit)
```
### tk_get_timeseries_summary
```{r}
# idx_date: First six columns, general summary
tk_get_timeseries_summary(idx_date)[,1:6]
```


```{r}
# idx_date: First six columns, differences summary
tk_get_timeseries_summary(idx_date)[,7:12]
```

```{r}
# idx_yearqtr: First six columns, general summary
tk_get_timeseries_summary(idx_yearqtr)[,1:6]
```

```{r}
# idx_yearqtr: Last six columns, difference summary
tk_get_timeseries_summary(idx_yearqtr)[,7:12]
```



# Making a Future Time Series Index
## Example 1: Remove Weekends and Last Two Weeks of Year

This is a simulated example that shows how to use the `tk_make_future_timeseries()` function. 

We’ll create a sequence of dates beginning in 2010, and extending through the first two weeks of 2013. 

The sequence is daily periodicity. We’ll split into training and test sets. The train set will be used to predict the future dates. The test set will be used for comparison.

```{r}
# Create sequence with omitted weekends and omitted last two weeks of each year
idx <- seq.Date(ymd("2010-01-01"), by = "day", length.out = 1110) %>%
    tk_get_timeseries_signature() %>%
    filter(!(wday.lbl) %in% c("Saturday", "Sunday")) %>%
    filter(!(week %in% c(51, 52, 53))) %>%
    tk_index()

idx_train <- idx[1:500]
idx_test  <- idx[501:length(idx)]
```

First we visualize the frequency of the simulated test index, which is what we will generate as a future time series. The `tk_get_timeseries_signature()` function retrieves the differences (frequency or “diff” column) of the test dataset. We can then plot using `ggplot()`. The frequency (diff) is in seconds. Since it’s daily periodicity, the base 86,400 seconds. Each small spike to 252,900 seconds is a weekend, and the large spike to 1,555,200 seconds is the missing two week period at the end of the year.

```{r}
idx_test %>%
    tk_get_timeseries_signature() %>%
    ggplot(aes(x = index, y = diff)) +
    geom_line(color = palette_light()[[1]]) +
    theme_tq() +
    labs(title = "Simple Test: Frequency of test set", 
         subtitle = "Missing weekends and missing last two weeks of year") +
    scale_y_continuous(limits = c(0, 2.2e6))
```

### Inspect weekdays

`inspect_weekdays` is useful for finding and removing missing dates that occur on a weekly, bi-weekly, tri-weekly or quad-weekly frequency (such as weekends or every other Friday off). When applied to the example, the algorithm finds and removes omitted weekends of the future data set.

```{r}
idx_train %>%
    tk_make_future_timeseries(n_future = 395, inspect_weekdays = TRUE) %>%
    tk_get_timeseries_signature() %>%
    ggplot(aes(x = index, y = diff)) +
    geom_line(color = palette_light()[[1]]) +
    theme_tq() +
    labs(title = "Simple Test: Frequency of predection with inspect_weekdays = T", 
         subtitle = "Catches missing weekends only") +
    scale_y_continuous(limits = c(0, 2.2e6))
```

### Inspect months

`inspect_months` is useful for finding and removing missing dates that occur on a monthly, quarterly, yearly frequency (such as last couple weeks of the year off). Applying to the example, the algorithm finds and removes omitted dates in the last two weeks of the year of the future data set.

```{r}
idx %>%
    tk_make_future_timeseries(n_future = 395, inspect_months = TRUE) %>%
    tk_get_timeseries_signature() %>%
    ggplot(aes(x = index, y = diff)) +
    geom_line(color = palette_light()[[1]]) +
    theme_tq() +
    labs(title = "Simple Test: Frequency of predection with inspect_months = T", 
         subtitle = "Catches missing last two weeks of year only") +
    scale_y_continuous(limits = c(0, 2.2e6))
```

### Inspect both weekdays and months

We can combine both inspect_weekdays and inspect_months to detect both sequences of missing dates. Applying both to the example removes the weekends and last two weeks of each year. However, there is one error made due to the interactions of the two algorithms.

```{r}
idx_future_wdays_and_months <- idx_train %>%
    tk_make_future_timeseries(n_future = 395, inspect_weekdays = T, inspect_months = T) 
idx_future_wdays_and_months %>%
    tk_get_timeseries_signature() %>%
    ggplot(aes(x = index, y = diff)) +
    geom_line(color = palette_light()[[1]]) +
    theme_tq() +
    labs(title = "Simple Test: Both inspect_weekdays = T and inspect_months = T", 
         subtitle = "For most part catches missing weekends and last two weeks of year") +
    scale_y_continuous(limits = c(0, 2.2e6))
```
### Reviewing errors

We can inspect further by analyzing both the Type I errors (errors that the algorithm removed incorrectly) and Type II errors (errors that the prediction that algorithm failed to remove).

#### Type I Errors

Errors that algorithm removed incorrectly. These errors are the most dangerous because users may not know which days were removed incorrectly. If known, they can be accounted for with `insert_values`. The easiest way to tell is by reviewing the frequency chart for larger than normal spikes.

```{r}
idx_test[!(idx_test %in% idx_future_wdays_and_months)]
```

#### Type II Errors

Errors that algorithm failed to remove. These errors are the eaisest to manage because typically the analyst generally knows which days should be removed. These errors can be addressed with `skip_values` provided prediction length is manageable.

```{r}
idx_future_wdays_and_months[!(idx_future_wdays_and_months %in% idx_test)]
```
### Using skip values and insert values

`insert_values` adds values to the future time series, and `skip_values` removes values from the future time series. The Type I errors (incorrectly removed observations) can be addressed with `insert_values`. The Type II errors (incorrectly kept observations) can be addressed with `skip_values`. We had one Type II error in the example, and we’ll correct with `skip_values`. Make sure the class of the value passed to `skip_values` matches the class of the time series index.

```{r}
idx_future_wdays_months_skip_vals <- idx_train %>%
    tk_make_future_timeseries(n_future = 395, 
                              inspect_weekdays = T, 
                              inspect_months   = T,
                              skip_values      = ymd("2012-12-01")) 
idx_future_wdays_months_skip_vals %>%
    tk_get_timeseries_signature() %>%
    ggplot(aes(x = index, y = diff)) +
    geom_line(color = palette_light()[[1]]) +
    theme_tq() +
    labs(title = "Simple Test: inspect_weekdays = T, inspect_months = T, skip_values", 
         subtitle = "Get the exact solution using skip values") +
    scale_y_continuous(limits = c(0, 2.2e6))
```


# Forecasting Using a Time Series Signature